MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
comment out nonessential code for testing

### Sex Check
```{r Reported_Gender_plots}
counts$gene_id = rownames(counts) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(counts) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  dplyr::mutate(value = log(value)) %>%
  dplyr::rename(`counts(log)`= value) %>% 
  dplyr::rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  dplyr::rename(`Reported Gender` = Reported_Gender)

##XIST and UTY expression 
#ENSG00000229807.11 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`)) 
p
```
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

## Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

@@

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('gene_id') %>%
  tidyr::gather(SampleID, logCPM, -gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
p
```

>>>>

<!-- ### Sex Check -->
<!-- ```{r Reported_Gender_plots} -->
<!-- counts$gene_id = rownames(counts)  -->
<!-- REPORTED.GENDER.COUNTS = GENE.GC.CONT %>%  -->
<!--   left_join(counts) %>% -->
<!--   dplyr::select(-one_of("percentage_gene_gc_content")) %>% -->
<!--   filter(chromosome_name == "X" |chromosome_name == "Y") %>%  -->
<!--   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>% -->
<!--   dplyr::mutate(value = log(value)) %>% -->
<!--   dplyr::rename(`counts(log)`= value) %>%  -->
<!--   dplyr::rename(SampleID = item) %>% -->
<!--   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>%  -->
<!--   dplyr::rename(`Reported Gender` = Reported_Gender)  -->

<!-- ##XIST and UTY expression  -->
<!-- #ENSG00000229807.11 and ENSG00000183878.15  -->
<!-- FILT <- REPORTED.GENDER.COUNTS %>%  -->
<!--   filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>%  -->
<!--   dplyr::select(-one_of("chromosome_name")) %>%  -->
<!--   tidyr::spread(key = gene_id, value = `counts(log)`) %>%  -->
<!--   mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>%  -->
<!--   mutate(UTY = as.numeric(`ENSG00000183878.15`)) -->

<!-- p = ggplot(FILT, aes (x= XIST, y = UTY))  -->
<!-- p = p + geom_point(aes(color=`Reported Gender`))  -->
<!-- p -->
<!-- ``` -->
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

<!-- ## Explore metatdata -->
<!-- ```{r data.explore, fig.width = 12, fig.height = 12} -->
<!-- my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5)) -->

<!-- # RIN -->
<!-- p = list() -->
<!-- p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot() -->
<!-- p[[1]] = p[[1]] + ggtitle('RIN') + my.theme -->

<!-- # Age of Death -->
<!-- p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot() -->
<!-- p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme -->

<!-- # PMI -->
<!-- p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot() -->
<!-- p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme -->

<!-- # Intronic Rate -->
<!-- p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot() -->
<!-- p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme -->

<!-- # IntergenicRate -->
<!-- p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot() -->
<!-- p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme -->

<!-- # Transcripts Detected -->
<!-- p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot() -->
<!-- p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme -->

<!-- # Mapped Reads -->
<!-- p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot() -->
<!-- p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme -->

<!-- # PercentAligned -->
<!-- p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot() -->
<!-- p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme -->

<!-- # rRNARate -->
<!-- p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot() -->
<!-- p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme -->

<!-- multiplot(plotlist = p, cols = 3) -->

<!-- # Institution -->
<!-- # vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES) -->

<!-- # Gender -->
<!-- # vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES) -->
<!-- ``` -->
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

@@

<!-- ### Co-expression distribution -->
<!-- Coexpression of genes  -->
<!-- ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5} -->
<!-- cr = cor(t(CQN.GENE_EXPRESSION$E)) -->
<!-- hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation') -->
<!-- ``` -->

<!-- ### Sample clustering -->
<!-- ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE} -->
<!-- # Find principal components of expression to plot -->
<!-- PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T) -->

<!-- # Plot first 2 PCs -->
<!-- plotdata <- data.frame(SampleID=rownames(PC$rotation),  -->
<!--                        PC1=PC$rotation[,1],  -->
<!--                        PC2=PC$rotation[,2]) -->

<!-- plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% -->
<!--   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') -->

<!-- p <- ggplot(plotdata, aes(x=PC1, y=PC2)) -->
<!-- p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath)) -->
<!-- p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.) -->
<!-- # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0) -->
<!-- p -->
<!-- ``` -->
<!-- Tree based clustering of samples -->
<!-- ```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'} -->
<!-- # Eucledian tree based analysis -->
<!-- COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")]) -->
<!-- COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0 -->

<!-- tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E))) -->
<!-- cols = WGCNA::labels2colors(COVARIATES.tmp); -->

<!-- WGCNA::plotDendroAndColors(tree,  -->
<!--                            colors = cols,  -->
<!--                            dendroLabels = FALSE,  -->
<!--                            abHeight = 0.80,  -->
<!--                            main = "Sample dendrogram", -->
<!--                            groupLabels = colnames(COVARIATES.tmp)) -->
<!-- ``` -->
<!-- ```{r temp, include = F} -->
<!-- dev.off() -->
<!-- gc() -->
<!-- ``` -->

<!-- ### Distribution of samples (log cpm) -->
<!-- ```{r lcpm.dist, cache=FALSE} -->
<!-- # Plot abberent distribution of logcpm counts -->
<!-- tmp1 = (CQN.GENE_EXPRESSION$E) %>% -->
<!--   rownameToFirstColumn('gene_id') %>% -->
<!--   tidyr::gather(SampleID, logCPM, -gene_id) %>% -->
<!--   left_join(COVARIATES %>% -->
<!--               rownameToFirstColumn('SampleID')) -->

<!-- p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density()  -->
<!-- p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free') -->
<!-- p -->
<!-- ``` -->
